Computing properties of independent sets by generic programming tensor networks \thanks{\funding{...}}

### 2022-01-08 paper/paper.tex
update revert table

This motivates us to think about what is the most general element type allowed in a tensor network contraction program.
We find that as long as the algebra of tensor elements forms a commutative semiring, the tensor network contraction result will be valid and be independent of the contraction order.
A commutative semiring is a ring with its multiplication operation being commutative and without an additive inverse.
To define a commutative semiring with the addition operation $\oplus$ and the multiplication operation $\odot$ on a set $R$, the following relations must hold for any arbitrary three elements $a, b, c \in R$.
\begin{align*}
(a \oplus b) \oplus c = a \oplus (b \oplus c) & \hspace{5em}\text{$\triangleright$ commutative monoid $\oplus$ with identity $\mymathbb{0}$}\\
a \oplus \mymathbb{0} = \mymathbb{0} \oplus a = a &\\
a \oplus b = b \oplus a &\\
&\\
(a \odot b) \odot c = a \odot (b \odot c)  &   \hspace{5em}\text{$\triangleright$ commutative monoid $\odot$ with identity $\mymathbb{1}$}\\
a \odot  \mymathbb{1} =  \mymathbb{1} \odot a = a &\\
a \odot b = b \odot a &\\
&\\
a \odot (b\oplus c) = a\odot b \oplus a\odot c  &  \hspace{5em}\text{$\triangleright$ left and right distributive}\\
(a\oplus b) \odot c = a\odot c \oplus b\odot c &\\
&\\
a \odot \mymathbb{0} = \mymathbb{0} \odot a = \mymathbb{0}
\end{align*}
We require the algebra being multiplicative commutative because the contraction order optimizer will change the multiplication order of elements, while we wish this contraction order does not change the contraction results.
In the following sections, we show how to compute a number of properties of independent sets by designing tensor element types as specific commutative semirings without changing the tensor network contraction program~\cite{Stepanov2014}.
The following diagram summarizes those properties that can be computed by various tensor element types.

% \begin{table}[t!]\centering
% \begin{minipage}{\columnwidth}
% \ra{1.3}
%         \begin{tabularx}{\textwidth}{sb}\toprule
%             \hline
%             \textbf{element type}     & \textbf{property/properties to compute} \\
%             {$\mathbbm{R}$}     & {the number of independent sets} \\
%             %{$\mathbbm{C}$}     & {independence polynomial (approximate)} \\
%             {$\text{GF}(p)$} (\Eq{eq:finitefield}) & {independence polynomial} \\
%             {T} (\Eq{eq:tropical})    & {MIS size} \\
%             {P1} (\Eq{eq:countingtropical})     & {MIS size and the number of MISs} \\
%             {P2} (\Eq{eq:max2poly})     & {the number of MISs and independent sets of size $\alpha(G)-1$} \\
%             {PN} (\Eq{eq:polynomial})     & {independence polynomial (direct approach)} \\
%             {P1+S1}     & {MIS size and one MIS} \\
%             {P1+SN} (\Eq{eq:countingtropicalset})    & {MIS size and all MISs} \\
%             \bottomrule
%         \end{tabularx}
%     \caption{Tensor element types and the independent set properties that can be computed using them.}\label{tbl:generictypes}
% \end{minipage}
% \end{table}

\centerline{\begin{tikzpicture}[]
    \scriptsize
    \node[ellipse,draw,minimum width=2cm, minimum height=1cm,inner sep=0pt] at (0, 2.5) (R) {$\mathbbm{R}$};
    % set and set sampler
    \node[ellipse,draw,minimum width=5cm, minimum height=3cm,inner sep=0pt] at (0, 0) (SN) {};
    \node[below of=1] at (SN) {Set};
    \node[ellipse,draw,minimum width=4cm, minimum height=1.5cm,inner sep=0pt] at (-0.5, 0) (S1) {\hspace{1.5cm}Bit string};
    \def\dx{-3};
    \node[ellipse,draw,minimum width=5cm, minimum height=3cm,inner sep=0pt] at (\dx, 0) (PN) {\hspace{-3.5cm}Polynomial};
    \node[ellipse,draw,minimum width=3.5cm, minimum height=1.2cm,inner sep=0pt] at (\dx, 0.7) (P1) {\hspace{-0.7cm}Highest order};
    \node[ellipse,draw,minimum width=3.5cm, minimum height=0.7cm,inner sep=0pt] at (\dx, -0.4) (P2) {\hspace{-0.7cm}2nd highest order};
    \node[ellipse,draw,minimum width=0.8cm, minimum height=0.3cm,inner sep=0pt] at (\dx, 1.0) (T) {};
    \node at (\dx, -1.2) {$\ldots$};
    \node[above of=2] at (T) (textT) {Tropical};
    \draw[black,-latex] (textT) -- (T);
\end{tikzpicture}}
\vspace{2em}
\centering{
\begin{minipage}{0.8\columnwidth}\scriptsize
\begin{tabularx}{\textwidth}{bb}
    \textbf{element type}     & \textbf{property to compute} \\
    {$\mathbbm{R}$}     & {the number of independent sets} \\
    {Polynomial} (\Eq{eq:polynomial})     & {independence polynomial} \\
    {Tropical (\Eq{eq:tropical})}    & {MIS size} \\
    {Polynomial truncated to highest order (\Eq{eq:countingtropical})}     & {MIS size and the number of MISs} \\
    {Polynomial truncated to 2nd highest order} (\Eq{eq:max2poly})     & {the number of MISs and independent sets of size $\alpha(G)-1$} \\
    {Set} (\Eq{eq:set})     & {enumeration of independent sets} \\
    {Bit string} (\Eq{eq:singleconfig})     & {one independent set} \\
    {Polynomial truncated to highest order combined with Bit string}     & {MIS size and one MIS} \\
    {Polynomial truncated to highest order combined with Set} (\Eq{eq:countingtropicalset})    & {MIS size and all MISs} \\
\end{tabularx}
\end{minipage}
}

@@

\section{Discussion and conclusion}
In this paper, we introduce an abstract algebra formalism to compute properties of independent sets.
The properties include the MIS size, number of independent set of a given size, and enumeration of independent sets of a give size.
For each property, we design a algebra being a commutative semiring, and map the property computation to the contraction of a tensor network with such element type.
%We derived the backward rule for tropical tensor network to bound the search of solution space.
%Although many of these properties are global, we can encode them to different tensor element types as commutative semirings.
We call this method generic programming tensor network, and its power is not limited to the independent set problem.
In \App{app:otherproblems}, we show how to map the maximal independent set problem, matching problem, $k$-colouring problem, max cut problem and set packing problem to tensor networks.
Moreover, since the independence polynomial is closely related to the matching polynomial~\cite{Levit2005}, the clique polynomial~\cite{Hoede1994}, and the vertex cover polynomial~\cite{Akbari2013},
our algorithm to compute the independence polynomial can also be used to compute these graph polynomials.
%Some properties of the independence polynomial, such as the uni-modality, log-concavity, and roots of the polynomial, are well studied in the literature for special classes of graphs~\cite{Levit2005}.
%We hope our algorithms to calculate the independence polynomial can also help further research in these directions.
We show some of the Julia language implementations in Appendix~\ref{sec:technical} and you will find it surprisingly short.
A complete implementation can be found in our Github repository~\cite{GraphTensorNetworks}.
%What we need to do is just defining two operations $\oplus$ and $\odot$ and two special elements $\mymathbb{0}$ and $\mymathbb{1}$.
%The style that we program is called generic programming,

>>>>

This motivates us to think about what is the most general element type allowed in a tensor network contraction program.
We find that as long as the algebra of tensor elements forms a commutative semiring, the tensor network contraction result will be valid and be independent of the contraction order.
A commutative semiring is a ring with its multiplication operation being commutative and without an additive inverse.
To define a commutative semiring with the addition operation $\oplus$ and the multiplication operation $\odot$ on a set $R$, the following relations must hold for any arbitrary three elements $a, b, c \in R$.
\begin{align*}
(a \oplus b) \oplus c = a \oplus (b \oplus c) & \hspace{5em}\text{$\triangleright$ commutative monoid $\oplus$ with identity $\mymathbb{0}$}\\
a \oplus \mymathbb{0} = \mymathbb{0} \oplus a = a &\\
a \oplus b = b \oplus a &\\
&\\
(a \odot b) \odot c = a \odot (b \odot c)  &   \hspace{5em}\text{$\triangleright$ commutative monoid $\odot$ with identity $\mymathbb{1}$}\\
a \odot  \mymathbb{1} =  \mymathbb{1} \odot a = a &\\
a \odot b = b \odot a &\\
&\\
a \odot (b\oplus c) = a\odot b \oplus a\odot c  &  \hspace{5em}\text{$\triangleright$ left and right distributive}\\
(a\oplus b) \odot c = a\odot c \oplus b\odot c &\\
&\\
a \odot \mymathbb{0} = \mymathbb{0} \odot a = \mymathbb{0}
\end{align*}
We require the algebra being multiplicative commutative because the contraction order optimizer will change the multiplication order of elements, while we wish this contraction order does not change the contraction results.
In the following sections, we show how to compute a number of properties of independent sets by designing tensor element types as specific commutative semirings without changing the tensor network contraction program~\cite{Stepanov2014}.
The Table~\ref{tbl:generictypes} summarizes those properties that can be computed by various tensor element types.

\begin{table}[t!]\centering
\begin{minipage}{\columnwidth}
\ra{1.3}
        \begin{tabularx}{\textwidth}{sb}\toprule
            \hline
   \textbf{element type}     & \textbf{property to compute} \\
   {$\mathbbm{R}$}     & {the number of independent sets} \\
   {Polynomial} (\Eq{eq:polynomial})     & {independence polynomial} \\
   {Tropical (\Eq{eq:tropical})}    & {MIS size} \\
   {Polynomial truncated to highest order (\Eq{eq:countingtropical})}     & {MIS size and the number of MISs} \\
   {Polynomial truncated to 2nd highest order} (\Eq{eq:max2poly})     & {the number of MISs and independent sets of size $\alpha(G)-1$} \\
   {Set} (\Eq{eq:set})     & {enumeration of independent sets} \\
   {Bit string} (\Eq{eq:singleconfig})     & {one independent set} \\
   {Polynomial truncated to highest order combined with Bit string}     & {MIS size and one MIS} \\
   {Polynomial truncated to highest order combined with Set} (\Eq{eq:countingtropicalset})    & {MIS size and all MISs} \\
            \bottomrule
        \end{tabularx}
    \caption{Tensor element types and the independent set properties that can be computed using them.}\label{tbl:generictypes}
\end{minipage}
\end{table}

% \vspace{2em}
% \centering{
% \begin{minipage}{0.8\columnwidth}\scriptsize
% \begin{tabularx}{\textwidth}{bb}
%     \textbf{element type}     & \textbf{property to compute} \\
%     {$\mathbbm{R}$}     & {the number of independent sets} \\
%     {Polynomial} (\Eq{eq:polynomial})     & {independence polynomial} \\
%     {Tropical (\Eq{eq:tropical})}    & {MIS size} \\
%     {Polynomial truncated to highest order (\Eq{eq:countingtropical})}     & {MIS size and the number of MISs} \\
%     {Polynomial truncated to 2nd highest order} (\Eq{eq:max2poly})     & {the number of MISs and independent sets of size $\alpha(G)-1$} \\
%     {Set} (\Eq{eq:set})     & {enumeration of independent sets} \\
%     {Bit string} (\Eq{eq:singleconfig})     & {one independent set} \\
%     {Polynomial truncated to highest order combined with Bit string}     & {MIS size and one MIS} \\
%     {Polynomial truncated to highest order combined with Set} (\Eq{eq:countingtropicalset})    & {MIS size and all MISs} \\
% \end{tabularx}
% \end{minipage}
%}

@@

\section{Discussion and conclusion}
In this paper, we introduce an abstract algebra formalism to compute properties of independent sets.
The properties include the MIS size, number of independent set of a given size, and enumeration of independent sets of a give size.
For each property, we design a algebra being a commutative semiring, and map the property computation to the contraction of a tensor network with such element type.
The data types introduced in the main text can be diagramatically represented as the following, where we use overlap to represent two algebras can be combined to create a new algebra.
\centerline{\begin{tikzpicture}[]
    \scriptsize
    \node[ellipse,draw,minimum width=2cm, minimum height=1cm,inner sep=0pt] at (0, 2.5) (R) {$\mathbbm{R}$};
    % set and set sampler
    \node[ellipse,draw,minimum width=5cm, minimum height=3cm,inner sep=0pt] at (0, 0) (SN) {};
    \node[below of=1] at (SN) {Set};
    \node[ellipse,draw,minimum width=4cm, minimum height=1.5cm,inner sep=0pt] at (-0.5, 0) (S1) {\hspace{1.5cm}Bit string};
    \def\dx{-3};
    \node[ellipse,draw,minimum width=5cm, minimum height=3cm,inner sep=0pt] at (\dx, 0) (PN) {\hspace{-3.5cm}Polynomial};
    \node[ellipse,draw,minimum width=3.5cm, minimum height=1.2cm,inner sep=0pt] at (\dx, 0.7) (P1) {\hspace{-0.7cm}Highest order};
    \node[ellipse,draw,minimum width=3.5cm, minimum height=0.7cm,inner sep=0pt] at (\dx, -0.4) (P2) {\hspace{-0.7cm}2nd highest order};
    \node[ellipse,draw,minimum width=0.8cm, minimum height=0.3cm,inner sep=0pt] at (\dx, 1.0) (T) {};
    \node at (\dx, -1.2) {$\ldots$};
    \node[above of=2] at (T) (textT) {Tropical};
    \draw[black,-latex] (textT) -- (T);
\end{tikzpicture}}
%We derived the backward rule for tropical tensor network to bound the search of solution space.
%Although many of these properties are global, we can encode them to different tensor element types as commutative semirings.
Except for real numbers, we have polynomials and truncated polynomials algebras. We combine them with set algebra and bit string algebra for set enumeration and sampling.
We call a tensor network with generic tensor element type a generic tensor network, and its power is not limited to solving the independent set properties.
In \App{app:otherproblems}, we show how to map the maximal independent set problem, matching problem, $k$-colouring problem, max cut problem and set packing problem to tensor networks.
Moreover, since the independence polynomial is closely related to the matching polynomial~\cite{Levit2005}, the clique polynomial~\cite{Hoede1994}, and the vertex cover polynomial~\cite{Akbari2013},
our algorithm to compute the independence polynomial can also be used to compute these graph polynomials.
%Some properties of the independence polynomial, such as the uni-modality, log-concavity, and roots of the polynomial, are well studied in the literature for special classes of graphs~\cite{Levit2005}.
%We hope our algorithms to calculate the independence polynomial can also help further research in these directions.
We show some of the Julia language implementations in Appendix~\ref{sec:technical} and you will find it surprisingly short.
A complete implementation can be found in our Github repository~\cite{GraphTensorNetworks}.
%What we need to do is just defining two operations $\oplus$ and $\odot$ and two special elements $\mymathbb{0}$ and $\mymathbb{1}$.
%The style that we program is called generic programming,
